---
title: "Snowshoe Hares in Bonanza Creek"
description: |
  Mini Report: Data Visualizations, T-Tests & Linear Regression Analysis
  
author:
  - name: Kathleen Cotti
    url: {}
date: 2021-02-20
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
    highlight: haddock
---

### Summary: 

This post includes a mini-report I completed for my Environmental Data Science course in Fall of 2020 at the Bren School. I learned how to customize `gg_beeswarm` plots and descriptive statistic tables with `kableExtra`. I also practiced running two-sample t-tests and linear regression analysis in R and describe the outcomes in this report. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Attach necessary packages:
library(tidyverse) 
library(janitor)
library(here)
library(effsize)
library(lubridate)
library(kableExtra)
library(ggpubr)
library(knitr)
library(broom)
library(car)
library(ggbeeswarm)
library(gapminder)
```

```{r, include = FALSE,warning = FALSE, message = FALSE}
#Read in the data using here: 
bonanza_hares_data <- read_csv(here("data", "bonanza_hares.csv"))
```

# A. Introduction: 
This report provides an exploratory overview of snowshoe hare physical data collected in the Bonanza Creek Experimental Forest in the Alaskan Tanana Valley from 1998 - 2012. Snowshoe hares are a keystone prey species in this region and the study aimed to describe fluctuations in snowshoe hare population density (Kielland et al. 2017). Specifically, the counts of trapped juvenile snowshoe hares was explored across the course of the study, juvenile snowshoe hare weight was compared between male and females across the three sampling sites, and the relationship(s) between hind foot length and juvenile weight were explored in this report. 

# B. Data & Analyses: 
Snowshoe hare weight and size measurements were collected from capture-recapture trapping and made available by Dr. Knut Kielland and colleagues at the Bonanza Creek LTER, a partnership between the University of Alaska Fairbanks and the US Forest Service and part of the US Long Term Ecological Research Network. The data contain observations for 3197 snowshoe hare trappings, collected over 14 years (1998-2012), in three sampling sites (bonrip or Bonanza Riparian, bonmat or Bonanza Mature, bonbs or Bonanza region with Mature Lowland Spruce). Changes in annual juvenile snowshoe trapping counts are displayed in a bar graph, and the distribution of juvenile snowshoes between sex and grid location is displayed using a beeswarm and boxplot. Following Exploratory Data Visualization, juvenile hare weights between male and female hares are compared by Welch's two-sample t-tests using a significance level of 0.05 throughout. Differences between groups are described by Cohen's *d* effect size. The relationship between hind foot length and weight is explored by a simple linear regression. Are analyses are in R version 4.0.2 using RStudio version 1.3.1056. 

# C. Exploratory Findings: 

### i) Annual Juvenile Hare Trap Counts and Descriptive Statistics:

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center", fig.cap = "**Figure 1:** Bar graph displaying counts of juvenile snowshoe hares trapped per year of the study (1998-2012). Sex and grid (sampling site) were not considered as variables here. No data is available for hare age in 1998, 2002, and 2009 therefore counts of juveniles trapped can not be included for these years. Data: Kielland et al. (2017)"}

#Create a new subset of data converting the dates in date format & with a new column for year:
bonanza_dates_converted <- bonanza_hares_data %>% 
  mutate(date_new = mdy(date)) %>%  
  mutate(year = year(date_new))

#Create a new subset filtering to keep only juvenile hares & group data by year and count:
juvenile_hare <- bonanza_dates_converted %>%  
  filter(age == "j")  %>% 
  count(age, year) 

#Create a bar graph of juvenile hare counts per year: 
ggplot(data = juvenile_hare, aes(x = year, y = n)) +
  geom_col(fill = "darkseagreen4") +
  labs(x = "Year", y = "Count", title = "Number of Juvenile Snowshoe Hares Trapped in Bonanza Creek \n Experimental Forest between 1998 and 2012") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_continuous(breaks = c(1998,1999,2000,2001,2002,2003,2004,2005,2006,2007,2008,2009,2010,2011,2012))

#Generate summary statistics for the juvenile hare data:
juvenile_summary <- juvenile_hare %>% 
  summarize(mean = mean(n), #Find the mean of the number of hares 
            median = median(n), #Find the median of the number of hares
            max =max(n), #Gives max
            min = min(n)) #Gives min

#Save summary stats to be used in analysis:
mean_hare <- juvenile_summary[1,1] #Stores mean as an object
median_hares <- juvenile_summary[1,2] #Stores median as an object
max_hares <- juvenile_summary[1,3] #Stores max as an object
min_hares <- juvenile_summary[1,4] #Stores min as an object
```

The minimum juvenile snowshoe hare trapping count in this region is `r min_hares`, and the maximum is `r max_hares`. According to the data available for this report, this minimum occurred in 2010 and the maximum occurred in 1999. The mean number of juvenile snowshoes trapped per year is `r mean_hare`. Since the maximum in 1999 is a large outlier, the mean is a skewed measure of central tendency. The median number of juvenile snowshoes trapped annually is `r median_hares`, and since this dataset is largely skewed by outliers this is a more accurate measure of central tendency. The general trend seen in the histogram is a decrease in the number of juvenile hares trapped from 1999 to 2012, with a large decrease between 1999 and 2001. Accounting for the number of traps placed per year and the time the traps were left out would help to standardize the juvenile count data further, since having a larger number of traps or traps being left out longer could correlate to increased numbers of trapped hares. It would be useful to have consistent data collection techniques, such as ages notated in each year. Since age data was not collected in 1998, 2002, and 2009 we can not see a trend across the entire data set for juvenile counts. 

## ii) Difference in Juvenile Hare Weight Between Male and Female across grid (sampling site) locations:

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align= "center", fig.cap = "**Figure 2:** Weight (g) observations for male, female, and NA (Undetermined) juvenile snowshoe hares in each of the three grid locations (bonbs, bonmat, and bonrip). Dark Green(Female), Golden Yellow(Male), and Gray (NA) points indicate individual observations for weight (g) of juvenile snowshoe hares. Box endpoints indicate the 25th and 75th percentile values; the black line and black point within the box indicate the median and mean value for each gender in each grid, respectively. Data: Kielland et al. (2017)"}

#Create a new subset of data containing weight, sex and location variables of juveniles:
j_hare_weight <-  bonanza_dates_converted %>% 
  filter(age == "j") %>%
  mutate(Sex = case_when(sex == "f" ~ "Female",  #Rename variables for clarity in graphical representation
                         sex == "m" ~ "Male")) %>% 
  mutate(Grid = case_when(grid == "bonbs" ~ "Bonanza", 
                          grid == "bonmat" ~ "Bonanza Mature", 
                          grid == "bonrip" ~ "Bonanza Riparian")) %>% 
  select(weight, Sex, Grid) 

#Generate summary statistics for the weight data grouped by sex and grid: 
j_hare_weight_summary <- j_hare_weight %>% 
  group_by(Sex, Grid) %>% 
  summarize(mean = mean(weight,na.rm=TRUE), 
            median = median(weight, na.rm=TRUE),
            sd = sd(weight, na.rm=TRUE),
            n = n()) 

#Create a beeswarm plot of weight distributed facet wrapped by grid with a boxplot on top displaying summary statistics: 
ggplot(data = j_hare_weight, 
       aes(x = Sex, 
           y = weight)) +
  geom_beeswarm(aes(color = Sex), alpha = 0.6) + 
  geom_boxplot(fill = NA, 
               width = 0.2, 
               outlier.color = NA) +
  stat_summary(fun=mean, 
               geom="point", 
               shape=20, 
               size=4, 
               color="black", 
               fill="black") +
  scale_color_manual(values = c("darkseagreen4","goldenrod2"), 
                     na.value = "gray10") +
  theme_grey() +
  labs(x = "Sex",
       y = "Weight (g)",
       title = "Weight of Male vs. Female Juvenille Snowshoe Hares in Each Grid",
       color = "Sex") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~Grid)
```

Weights were compared across all three grid sites and between sex of the juvenile snowshoe hare population. On average male hares in each grid location have larger mean and median body weights in each of the three grid zones than female hares and hares with undetermined sex. The average weights of both male and female hares was largest in the Bonanza grid than the other grid locations. All distributions appear relatively normally and symmetrically distributed. The Bonanza grid has a significant male outlier causing the mean and median to differ, as the mean is skewed negatively. The Bonanza grid female distribution tail shows slight negative skew but with no significant outliers and similar mean and median values. The Bonanza Mature grid distributions show no significant outliers, but the female distribution mean is skewed negatively. The Bonanza Riparian distribution of Females has a significant positive outlier, but the mean and median values are relatively similar. Overall, the data appears relatively normally distributed between sex and grid and we can continue with the data analysis. 


```{r,include = FALSE}
#Exploratory Graphs - Histograms & QQ Plots to explore assumptions before running t-test
ggplot(data = j_hare_weight, aes(x = weight)) +
  geom_histogram(bins = 8,
                 aes(fill = Grid)) +
  facet_wrap(~Grid) 

ggplot(data = j_hare_weight, aes(sample = weight)) +
  geom_qq()+
  facet_wrap(~Grid)

ggplot(data = j_hare_weight, aes(x = weight)) +
  geom_histogram(bins = 8,
                 aes(fill = Sex)) +
  facet_wrap(~Sex)

ggplot(data = j_hare_weight, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~Sex)
```


## iii.) Difference in Weight Between Male and Female Juvenile Snowshoe Hares:

As a starting point, the distribution of male and female juvenile hare weights were explored using both histograms and quantile-quantile plots. The distributions are relatively normal, and both our sample  allowing us to move forward with our t-test.  

```{r, echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}

#Make a subset of data for this comparison:
j_gender_comparison <- j_hare_weight %>%  
  select(!Grid) %>% 
  filter(Sex %in% c("Male", "Female")) 

#Make a summary table:
j_gender_summary <- j_gender_comparison %>%  
  group_by(Sex) %>% 
  summarize(Mean = mean(weight, na.rm=TRUE),
            sd = sd(weight, na.rm=TRUE),
            n =n()) 

#Rename the Columns in the Table for clarity in final representation: 
colnames(j_gender_summary) 
names(j_gender_summary)[names(j_gender_summary) == "Mean"] <- "Mean Weight (g)"
names(j_gender_summary)[names(j_gender_summary) == "sd"] <- "Standard Deviation Weight (g)"
names(j_gender_summary)[names(j_gender_summary) == "n"] <- "Sample Size"

```

**Table 1**: Descriptive Statistics (Mean,Standard Deviation, and Sample Size) of Male and Female Juvenile Snowshoe Hares from all grids. Data: Kielland et al. (2017)

```{r, message = FALSE, echo = FALSE, warning = FALSE}
#Create an organized Table in the Knitted Doc:
kbl(j_gender_summary) %>% 
  kable_material("hover", full_width = F)

# Get the Means and Standard Deviations for each stored as objects:
mean_f <- j_gender_summary[1,2] 
mean_m <- j_gender_summary[2,2]
sd_f <- j_gender_summary[1,3]
sd_m <- j_gender_summary[2,3]

#Pull vectors for weight for male & female juvenile hares:
j_female <- j_gender_comparison %>% 
  filter(Sex == "Female") %>% 
  filter(weight > 0) %>% 
  pull(weight)

j_male<- j_gender_comparison %>% 
  filter(Sex == "Male") %>% 
  filter(weight > 0) %>% 
  pull(weight)

#Run the T-test: 
gender_ttest <- t.test(j_female, j_male)
gender_ttest_tidy <- tidy(gender_ttest)

#Get the Effect Size:
gender_effsize <- cohen.d(j_female, j_male)

```

On average, juvenile male snowshoe hairs have larger body weights than juvenile female snowshoes (`r round(mean_m,2)` $\pm$ `r round(sd_m, 2)` and `r round(mean_f,2)` $\pm$ `r round(sd_f, 2)` mm, respectively; mean $\pm$ 1 standard deviation). While the absolute difference in means is `r round(mean_m, 2) - round(mean_f, 2)` mm (a `r round((mean_m - mean_f)/((mean_m + mean_f)/2)*100, 2)`% difference). This difference in means is significant, as the p-value is below our established alpha value of 0.05 (Welch's two-sample t-test: t(`r round(gender_ttest_tidy$parameter,2)`) = `r round(gender_ttest_tidy$statistic,2)`, p < 0.001), meaning we would reject the null hypothesis and assume that the samples were drawn from populations with different means. The effect size is small (Cohen's *d* = `r round(gender_effsize$estimate,2)`) meaning there is a difference in the means but it is difficult to notice. 

## iv) Relationship between juvenille weight & hind foot length:

As a starting point, the relationship between juvenile weight and hind foot length was explored across the sexes (ie. ignoring sex) as seen in Figure 3. Further analysis is needed to compare the relationship between juvenile weight and hind foot length within each sex, and should include sex as a variable.

```{r, echo = FALSE, message = FALSE, warning = FALSE, include =FALSE}
#Create a new subset of the data for this exploration:
hare_foot <- bonanza_dates_converted %>% 
  filter(age == "j") %>% 
  select(hindft, weight,sex) %>% 
  filter(weight > 0) %>% #Filters out NA data points from the weight and hindft columns
  filter(hindft > 0 )
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.cap= "**Figure 3:** Relationship between juvenile hare foot length (mm) and body weight (g).Points indicate individual snowshoe hare measurements."}
#Create a scatterplot to explore the relationship between variables: 
ggplot(data = hare_foot, 
       aes(x = hindft,
           y = weight)) +
         geom_point(color = "darkseagreen4",
                    show.legend = FALSE,
                    size = 2.5,
                    alpha = 0.7) +
  labs(x = "Hind Foot Length (mm)",
       y = "Body Weight (g)", 
       title = "Relationship between Juvenile Hare Body Weight and Hind Foot Length") +
  theme_gray() +
  theme(plot.title = element_text(hjust = 0.5))

```


```{r, echo = FALSE, message = FALSE, warning = FALSE, include = FALSE }
#Do Linear Regression:
hare_lm <- lm(weight ~ hindft, data = hare_foot)

# See model diagnostics:
plot(hare_lm)

# Get Pearson's r correlation
hare_cor <- cor.test(hare_foot$hindft, hare_foot$weight) 

# Tidy version of correlation output:
hare_lm_tidy <- broom::tidy(hare_lm)
hare_lm_glance <- glance(hare_lm)
hare_cor_tidy <- tidy(hare_cor)
```


```{r,echo= FALSE, message = FALSE, warning = FALSE}
#Plot the Linear Regression: 
ggplot(data = hare_foot, aes(x = hindft, y= weight)) +
  geom_point(size = 1.5) +
  geom_smooth(method = "lm",
              color = "darkseagreen4",
              size = 2,
              fill = "goldenrod2",
              alpha = 0.2) +
  labs(x = "Hind Foot Length (mm)", y = "Weight (g)", title = "Linear Regression Model Describing how Snowshoe Hare Weight \n Changes with respect to Hind Foot Length") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggpubr::stat_regline_equation(label.x = 65, label.y = 1500) 
```

**Figure 4:** Relationship between hindfoot length (mm) and body weight (g) for all juvenile snowshoe hares. Points indicate individual hare measurements. Linear model summary: $\beta$~1~ = `r round(hare_lm_tidy$estimate[2], 2)` g mm^-1^, p < 0.001, R^2^ = `r round(hare_lm_glance$r.squared,2)`, Pearson's *r* = `r round(hare_cor_tidy$estimate,2)`).

The slope of the regression equation is $\beta$ = `r round(hare_lm_tidy$estimate[2], 2)`, meaning for every one unit increase in hind foot length (mm) we expect an average increase in hare weight of $\beta$ = `r round(hare_lm_tidy$estimate[2], 2)` grams. The R^2^ value is `r round(hare_lm_glance$r.squared,2)` meaning 30% of the variation in juvenile hare weight can be explained by the variation in hind foot length. The Pearson's *r* is *r* = `r round(hare_cor_tidy$estimate,2)` which indicates there is a medium level of correlation between the variables. These results indicate there is positive correlation between the variables; however, since only 30% of the variation in weight can be explained by the hind foot length a multiple regression with more explanatory variables would be needed to explain the variation in juvenile hare weights, such as possibly accounting for the sex, age, or length of the hare. 

Diagnostic plots (not included) show that this model does not align with the assumptions for linear regression. The QQ plot is not linear throughout indicating the residuals are not normally distributed. The cooks distance test shows values outside the red dotted line indicating the possibility of outliers that can skew the model. The residuals appear heteroskedastic and the spread appears to be increasing across the model. There is an obvious pattern that the residuals are not normally distributed across the predicted trendline; therefore, linear regression was not the best model for this data and multiple regression should be considered. 

# D. Summary: 
Exploratory data analysis reveals the following initial findings:

- The number of juvenile snowshoe hare trapping counts decreased over the course of the study (1998-2012).

- Male juvenile snowshoe hares have larger mean and median body weights than female juveniles (the difference is significant and the effect size is small).

- Hind Foot Length is not perfectly linearly related with hare weight across all juvenile hares in the dataset; however, there is positive correlation between the variables (r* = `r round(hare_cor_tidy$estimate,2)`) , and a portion of the variation in body weight can be explained by the variation in hind foot length  (R^2^ =`r round(hare_lm_glance$r.squared,2)`).

# E. Citations:
Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. [URL Link](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-bnz.55.22)


